### 2015-03-28 markdown/manuscript.tex
fixed a typo

But insofar as GP \emph{surprises us}---and that is its sole strength over more predictable and manageable frameworks---we will inevitably see a good fraction of those surprises as \emph{disappointments} rather than opportunities to adapt ourselves. I'd like to suggest we learn new ways to \emph{accommodate} those disappointments, and stop trying to make them go away.

@@

\begin{enumerate}
\item Add a little test
\item Run all tests and fail
\item Make a little change
\item Run the tests and succeed
\item Refactor to remove duplication
\end{enumerate}

Each stage offers a stumbling block for an experienced programmer, but the most salient for us now is the overarching flow of implementation (or ``design'') that it imposes: Each cycle, and also the overarching design process, begins with the choice of \emph{which little test should next be added}; each cycle ends with a rigorous process of refactoring, not just of the new code but of the \emph{entire cumulative codebase} produced over all iterations of this cycle. The middle three steps---implementing a \emph{single} failing test and modifying the codebase \emph{by just enough} so that all tests pass---feel when one is working through them as if they could be automated easily. The \emph{mindfulness} of the process lives in the choice of next steps and (though somewhat less so) of standard refactoring operations.

Braithwaite's exercise does an interesting thing to surface the formal rigor of this approach. In it, the participants (willing, of course, because the exercise is a \emph{kata} or ``refresher'' for experienced software developers to hone their skills) are asked to implement a nominally simple project like the game of Tic-Tac-Toe, given an \emph{ordered} list of features to implement and the artificial restriction that they must go farther than normal TDD practice asks. Rather than producing a suite of tests and a self-contained codebase, they are forced to use \emph{only} refactoring of code added to tests to produce their eventual ``codebase''. In other words, no code can be ``produced'' until \emph{duplication in code added multiple passing tests} provides a warrant for refactoring it out. Further, a facilitator patrols ongoing work and deletes \emph{any and all code not called for by a pre-existing failing test}.

Words like ``irritating'' and ``annoying'' crop up in participants' accounts of  this onerous backtracking deletion the first few times it happens\ldots{} as one might imagine. But as Gojko Adzik emphasizes in his descriptions of workshops he's run, the results of ``design'' of even simple algorithms in this artificial amplified setting seems much more \emph{open-ended} than it would if the software were built within the typical framing of habits and assumptions that an experienced programmer carries along with her to any project.

@@

\section{As if we \emph{really} meant it}\hypertarget{as-if-we-really-meant-it}{}\label{as-if-we-really-meant-it}

@@

\subsubsection{``as if we really meant it''}\hypertarget{as-if-we-really-meant-it-1}{}\label{as-if-we-really-meant-it-1}

>>>>

But insofar as GP \emph{surprises us}, and since that is its sole strength over more predictable and manageable frameworks, we must inevitably see a good fraction of those surprises at least in part as \emph{disappointments} rather than encouraging opportunities to change our plans.  Let's learn new ways to accommodate those disappointments, and stop trying to make them go away.

@@

\begin{enumerate}
\item Add a little (failing) test which exercises the next behavior you want to build into your codebase
\item Run all tests, expecting only the newest to fail
\item Make the minimal change to your codebase that permits the new test to pass
\item Run all tests, expecting them all to succeed
\item Refactor codebase to remove duplication
\end{enumerate}

Each stage offers a stumbling block for an experienced programmer, but the most salient for us now is the iterative flow of implementation (or ``design'') that it imposes: Each cycle begins with a choice of \emph{which little test should next be added}; each cycle ends with a rigorous process of refactoring, not just of the new code but of the \emph{entire cumulative codebase} produced so far. The middle three steps---implementing a \emph{single} failing test and modifying the codebase \emph{by just enough} so that all tests pass---feel when one is working through them as if they could be automated easily. The \emph{mindfulness} of the process lives in the choice of next steps and (though somewhat less so) of standard refactoring operations.

Braithwaite's exercise does an interesting thing to surface the formal rigor of this approach. In it, the participants (willing, of course, because the exercise is a \emph{kata} or ``refresher'' for experienced software developers to hone their skills) are asked to implement a nominally simple project like the game of Tic-Tac-Toe, given an \emph{ordered} list of features to implement and the artificial restriction that they must go farther than normal TDD practice asks. Rather than producing a suite of tests and a self-contained codebase, they are forced to use \emph{only} refactoring of code added to tests to produce their eventual ``codebase''. In other words, no code can be ``produced'' until the ``smell'' of duplication in the code added to multiple passing tests \emph{provides a warrant} for refactoring it out.

Further, a facilitator patrols ongoing work and deletes \emph{any and all code not called for by a pre-existing failing test}. Words like ``irritating'' and ``annoying'' crop up in participants' accounts of  this onerous backtracking deletion the first few times it happens, as one might imagine. But as Gojko Adzik emphasizes in his descriptions of workshops, the resulting designs for even simple algorithms in this artificially amplified setting seems much more \emph{open-ended} than it would if the software were built under the typical norms and habits an experienced programmer uses in normal conditions.

@@

\section{GP as if we meant it}\hypertarget{gp-as-if-we-meant-it}{}\label{gp-as-if-we-meant-it}

@@

\subsubsection{``as if we really meant it''}\hypertarget{as-if-we-really-meant-it}{}\label{as-if-we-really-meant-it}
